%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion                    %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

This chapter interprets the results presented in Chapter 5 and situates them within their academic, practical, and regulatory context. It first discusses the meaning of the sector diagnostic and the resulting maturity profile, then articulates the theoretical and managerial implications of the framework and the DataPilot tool, compares the contribution with the existing state of practice, makes explicit the rigor of the pre-development work on which the artifacts rest, and finally examines the threats to the validity of the work and the measures taken to mitigate them.

\subsection{Interpretation of the Key Findings}

The central finding of the study is that the data maturity of the Tunisian banking sector is concentrated at the emerging level, the second of the five maturity levels. This is not a marginal observation: it means that the sector as a whole has moved beyond the purely ad hoc stage but has not yet reached the threshold of formalization, standardization, and repeatability that defines the third level. The worked example of a Global Maturity Index (GMI) of 2.19 out of 5, equivalent to 44\%, is representative of this condition. An institution at this level typically possesses isolated good practices but lacks the documented, approved, and consistently applied processes that would make those practices durable and auditable.

The diagnostic data explains why the sector is positioned at this level rather than higher. The governance dimension, which carries the largest weight at 25\%, is also the weakest: only 30\% of respondents report a formally approved data strategy, so roughly seven in ten respondents work without a board-level data mandate. Because governance is the foundation on which the other dimensions rest, a deficit here propagates: without an approved strategy, data ownership is rarely formalized, quality controls are not mandated, and investment in architecture and skills is not prioritized. The data quality finding reported in Chapter 5 reinforces this interpretation. That a majority of respondents verify data before use is, on its surface, a sign of diligence, but it is more accurately read as a symptom of distrust: professionals verify because they cannot rely on the data, which indicates the absence of systematic, automated quality controls upstream. The architecture findings, where access is widely described as slow and centralization as only partial, show that even where governance intent exists, the technical substrate does not yet support it. Finally, the analytics finding, in which data-driven decision making remains a minority practice against a majority still relying on intuition, quantifies the business cost of the maturity deficit: the sector is not yet extracting decision value from its data at scale. The underlying dimension-by-dimension figures are reported in full in Chapter 5 and are not repeated here.

Taken together, these findings describe a coherent and self reinforcing pattern rather than a set of isolated weaknesses. This is an important interpretive point, because it implies that piecemeal remediation, addressing one dimension in isolation, is unlikely to raise the overall maturity. The weighted and multidimensional structure of the framework is a direct response to this reality: it makes the interdependence of the dimensions explicit and discourages the common error of investing in analytics tools while leaving the governance and quality foundations unaddressed.

The choice of weights is itself an interpretive statement about this interdependence. Governance receives the largest share at 25\% precisely because it is causally upstream of the other four dimensions: a strategy, an ownership model, and a compliance mandate determine whether quality controls are funded, whether architecture is consolidated, and whether analytics investment is directed at genuine business questions. Data quality, architecture, and analytics each carry 20\%, reflecting that they are the operational layers through which governance intent is realized and through which decision value is ultimately produced. Skills and culture carries 15\%, the smallest share, not because it is unimportant but because it is assessed only through proxy indicators and because its effect on maturity is mediated by the other four dimensions rather than expressed directly in auditable artifacts. The weighting therefore encodes a theory of how maturity is built, and the diagnostic profile, in which the heavily weighted governance dimension is also the weakest, explains why the aggregate index settles at the emerging level rather than higher.

It is also worth reading the profile through the lens of the maturity band definitions themselves. The emerging level is defined by the presence of isolated good practices that are not yet documented, approved, and consistently applied. Every headline finding fits this definition. Verification of data before use is a good practice, but performed manually and reactively rather than enforced by automated upstream controls. Partial centralization is progress, but it is progress that has not been completed or standardized. A minority of data driven decisions shows that the capability exists, but it has not been generalized into the default mode of decision making. The diagnostic thus does not describe a sector that is failing; it describes a sector that has begun the right activities but has not yet institutionalized them, which is exactly the transition from the second to the third maturity level that the framework is designed to measure and to guide.

\subsection{The Perception versus Reality Gap}

A finding that emerged during the design of the scoring methodology, and that has implications beyond this project, is the perception versus reality gap. Institutions consistently tend to rate their own maturity above the level their documentary evidence can support. A bank may, for example, declare that it has a data strategy, when what exists is an informal and undated set of intentions that has never been approved at board level. The evidence requirement and the evidence cap were designed specifically to neutralize this bias: any indicator scored above two out of five that is not backed by a verifiable artifact is automatically capped at two. This rule transforms the assessment from a declarative exercise, in which a bank rates itself, into an evidence based one, in which a bank must demonstrate. The methodological consequence is that the GMI produced by DataPilot is not a measure of how mature an institution believes it is, but of how mature it can prove it is, which is precisely the standard of proof that a regulatory audit expects.

The gap is not merely a nuisance to be corrected; it is diagnostically informative in its own right. The distance between the score an institution would assign itself and the score its evidence supports is a direct measure of the formalization deficit that separates the emerging level from the defined level. Where the two coincide, the institution has documented its practice; where they diverge, the institution possesses a capability in practice but has not yet made it durable, transferable, and auditable. Because the cap is applied deterministically at the level of each individual indicator rather than as a global adjustment, the assessment localizes this deficit precisely, showing not only that formalization is incomplete but exactly which controls, owners, or policies lack the artifact that would make them defensible. The tool therefore turns a well known cognitive bias in self assessment into a targeted list of documentation gaps, each of which corresponds to a concrete remediation action in the improvement roadmap.

This design also aligns the instrument with the logic of regulatory supervision. A BCT inspection does not accept a declaration that a control exists; it requires the artifact, the dated policy, the approval record, the lineage documentation, or the audit trail. By making evidence the precondition for any score above two, the framework is designed to align with the evidentiary standard that a regulator applies, so that a favorable DataPilot result reflects a state of readiness for the kind of scrutiny an inspection involves. The tool does not, however, substitute for a supervisory judgment: only the regulator can pronounce on compliance, and the instrument makes no claim to predict or replace that judgment. The alignment between the internal maturity measure and the external supervisory standard is a deliberate design goal rather than a coincidence.

\subsection{Theoretical and Academic Contributions}

The study makes three contributions to the body of knowledge on data maturity assessment. The first is the demonstration that no single established framework is sufficient for a regulated and sector specific context, and that a structured, criteria based selection and hybridization of multiple frameworks produces a more complete instrument than any of its sources. The screening of fifteen candidate frameworks against five selection criteria, and the explicit mapping of what was retained and what was excluded from each of the five selected frameworks, provides a replicable method for constructing hybrid maturity models in other regulated sectors.

The second contribution is the formalization of an auditable scoring methodology. Much of the existing maturity literature stops at descriptive level definitions and does not specify how indicator responses aggregate into a defensible global score, nor how to prevent score inflation. By defining a transparent aggregation chain, sub-dimension average, dimension average, and weighted global sum, and by attaching to it a set of explicit business rules, the evidence cap, the skip logic with a twenty percent ceiling, and the non-skippable status of regulatory indicators, the methodology closes the gap between a qualitative model and a quantitative, traceable measurement instrument.

The third contribution is the explicit alignment of a data maturity framework with a specific national regulatory text, the BCT Circulaire N\textdegree{}2025-08, in combination with the international BCBS 239 standard. The thirteen regulatory indicators embedded in the framework, and their non-skippable status, make regulatory compliance a measurable component of maturity rather than a separate exercise. This integration of a maturity model with the author's mapping to a national regulation, the BCT circular, together with the internationally recognized BCBS 239 principles, is, to the knowledge of this study, not present in the publicly available frameworks reviewed. The distribution of these thirteen indicators is itself meaningful: seven fall within governance, three within data quality, two within architecture, one within analytics, and none within skills and culture. This concentration in governance and quality expresses, at the level of the instrument, the regulator's own priority ordering, in which accountability, traceability, and the integrity of reported data precede tooling and culture. The framework thus does not treat compliance as an external checklist bolted onto a maturity model; it embeds the regulatory requirements at the exact dimensions where they logically belong, so that a compliance failure and a maturity weakness point to the same underlying gap.

A further theoretical implication follows from these three contributions taken together. The prevailing treatment of data maturity in the literature separates three concerns that this work unifies: the descriptive question of what capabilities define maturity, the measurement question of how to score them defensibly, and the regulatory question of whether the institution meets its obligations. By binding a hybridized capability model to a deterministic and auditable scoring method and to a specific regulatory mapping, the framework demonstrates that these three concerns can be addressed within a single coherent instrument rather than by three disjoint exercises. This unification is the study's principal theoretical claim, and it is what allows a single number, the GMI, to carry simultaneously a diagnostic, a measurement, and a compliance meaning.

\subsection{Practical and Managerial Implications}

For bank management, the framework and the DataPilot tool convert an abstract and frequently avoided question, how mature is our data capability, into a concrete, quantified, and trackable figure. The implications are threefold. First, the GMI provides a single number that can be reported to the board and tracked over successive assessments, which turns data maturity into a managed objective rather than an aspiration. Because the index is deterministic and evidence based, a change in the figure between two assessments reflects a real change in documented capability rather than a shift in respondent optimism, which makes it a defensible key performance indicator for a data transformation program. Second, the gap analysis and the three phase improvement roadmap translate the diagnosis into a prioritized action plan, in which each dimension is assigned an action band according to its score, namely a critical band for dimensions below two, a high priority band for dimensions below three, and a moderate band at or above three. The recommendations attached to each band are concrete and dimension specific, so that a governance score in the critical band surfaces actions such as approving a data strategy at board level and appointing an accountable executive, while the same dimension in a higher band surfaces optimization actions such as an annual KPI linked strategy review. The resulting sequence runs from the critical governance and compliance gaps that must be addressed first to the optimization work that follows once the foundations are in place. This sequencing directly reflects the interdependence of the dimensions discussed above, because remediation that begins with the causally upstream governance layer is the only ordering that allows the downstream investments in quality, architecture, and analytics to hold. Third, the compliance view and the exportable evidence package give management a defensible artifact to present to the regulator, reducing the cost and uncertainty of preparing for a BCT inspection, and turning what is usually a reactive scramble into a standing, continuously maintained record.

For EY Advisory Tunisia, the framework and the tool constitute a reusable consulting asset. A standardized, evidence based assessment instrument allows the firm to deliver maturity diagnostics with consistent rigor across engagements, to benchmark institutions against one another using a common scale, and to ground its improvement recommendations in a transparent and auditable methodology. The tool reduces the manual effort of each assessment and improves the comparability of the results. Just as importantly, it de-risks the advisory relationship: because the scoring rules are explicit and the evidence cap prevents inflation, the firm can defend each score to a client and to a regulator on the same objective basis, which protects the credibility of the recommendations that follow from it. The reusability of the instrument also means that the intellectual effort invested in framework design is amortized across many engagements rather than repeated for each, and that consultants can concentrate their expertise on interpretation and remediation rather than on reconstructing an assessment method every time.

The delivery of the tool as a multi-tenant platform, in the fourth iteration, extends these implications from the instrument itself to the way it is operated. The platform turns the instrument from a deliverable handed over at the end of an engagement into a service that EY operates across engagements: EY maintains the EY master template centrally, so that every engagement starts from the same calibrated instrument, while each bank receives its own copy of the questionnaire, which its Admin may tailor to local vocabulary and emphasis without any ability to alter the scoring rules, the evidence cap, the skip ceilings, or the compliance thresholds; because every copy starts from the canonical master template and EY reviews submissions across banks, local tailoring remains visible and results remain methodologically comparable across institutions. Group assessments distribute the questionnaire to the departments that actually hold the evidence, with each dimension assigned to the department best placed to answer it, replacing a single proxy respondent with accountable contributions whose authorship is recorded, and with a coordinator consolidating the work into one auditable submission per bank. The submissions that accumulate across banks form, over time, the comparable evidence base that the sector level benchmarking discussed below presupposes. What protects each bank in this shared setting is no longer the local custody of the data but tenant isolation enforced in the database itself, under which every row is confined to its own bank, together with invite-only access and the immutability of submitted results.

For the regulator and the sector as a whole, a common assessment standard creates the possibility of sector level benchmarking. If multiple institutions are assessed with the same framework and the same scoring rules, their results become comparable, which enables the identification of systemic weaknesses, such as the governance deficit revealed by the diagnostic, and supports the design of sector wide remediation policy. Because thirteen of the indicators are drawn directly from the regulatory texts, an aggregate view across institutions would also give the supervisor a sector level reading of compliance with the circular, distinguishing requirements that are broadly met from those, such as the formal approval of a data strategy, where the whole sector lags. A shared and auditable instrument thus offers the regulator a more precise instrument of supervision than institution by institution declarations, and offers the sector a common language in which to discuss and compare its progress.

\subsection{Comparison with the Existing State of Practice}

The contribution of this work is best understood by comparison with the frameworks from which it draws. The established frameworks each excel on a single axis: DAMA-DMBOK on structural breadth, DCAM on regulatory alignment for finance, CMMI-DMM on the maturity scale, TDWI on evaluation mechanics, and Gartner on executive communication. None, however, delivers all of these simultaneously, and none is operationalized as a usable tool aligned with the Tunisian regulatory environment. The framework designed in this project does not claim to replace these references; it composes them. Its novelty lies in the integration: a structurally complete model that scores in an auditable way, communicates to executives, and maps to a national regulation, delivered through a governed platform that a bank operates with its own accounts and that an advisor can administer across institutions. This combination addresses the three research gaps identified in Chapter 2, the absence of a sector specific and regulation aligned framework, the absence of an auditable scoring method, and the absence of a usable tool, each of which the existing state of practice leaves open.

The distinction between composition and replacement deserves emphasis, because it defines both the value and the honest limits of the contribution. Each source framework was retained for the axis on which it is strongest and set aside where another source was stronger, so the hybrid inherits the structural completeness of a broad reference model, the finance and regulatory orientation of a sector model, the calibrated five level scale of a maturity model, the scoring discipline of an evaluation oriented model, and the executive readability of a communication oriented model. What the hybrid adds beyond this composition is the connective tissue that none of the sources provides on its own: the deterministic aggregation chain that turns level definitions into a single number, the evidence cap that prevents that number from being inflated, and the author's regulatory mapping that ties thirteen indicators to a specific national text. The result is not a larger checklist but a different kind of instrument, one that is at once descriptive, measurable, auditable, and regulatorily grounded.

Against the state of practice in Tunisian banking specifically, the comparison is sharper still. In the absence of a shared instrument, institutions have assessed their data capability, when they have assessed it at all, through informal and non comparable self appraisals, and preparation for regulatory scrutiny has typically been a reactive, engagement specific effort. The framework replaces this fragmented practice with a single, reproducible, and evidence based standard that produces results comparable across institutions and across time. This is the practical sense in which the contribution advances the state of practice: not by proposing a novel definition of maturity, which the literature already supplies in abundance, but by making maturity in this regulated context something that can be measured the same way twice, defended to a supervisor, and acted upon through a prioritized roadmap.

\subsection{Rigor of the Pre-Development Work}

The credibility of the artifacts discussed above rests to a large extent on work that was completed before a single line of the tool was written. It is therefore worth making explicit why the conceptual phases of the project, the benchmark, the framework design, and the scoring methodology, meet the standard of rigor expected of a research contribution and of a professional consulting engagement, rather than being a thin preamble to software development.

The first element of this rigor is the breadth of the benchmark. Rather than adopting the first plausible reference model, the study screened fifteen international frameworks against five selection criteria, namely international credibility, banking relevance, scoring logic, complementarity, and BCT and BCBS 239 alignment, each framework being scored out of fifteen in a full comparison matrix. Every acceptance and every rejection was justified with a documented reason: only five frameworks were retained, and the ten rejections are not silent omissions but explicit, argued decisions, ranging from vendor lock-in and proprietary non-referenceable content to a scope too narrow to cover the five dimensions. Notably, the selection was not a mechanical score threshold; a proprietary model with a competitive score was still rejected because it could not be referenced academically, which shows that the selection rule was principled rather than arithmetic.

The second element is that the resulting framework is a genuine synthesis rather than an adoption. Each of the five selected frameworks was assigned a defined role in the hybrid, and for each of them the parts that were deliberately excluded are named with the reason for their exclusion. The model that emerges was engineered from complementary components, not borrowed from a single source and relabeled, and the traceability from every structural choice back to its source framework and its exclusion decisions is preserved in the benchmark documentation.

The third element is that the design is evidence driven. Each of the four directly observable dimensions is justified by a specific finding of the diagnostic survey of 44 banking professionals, and the fifth is justified by the survey's inability to observe it directly, which is itself the finding that motivated the proxy-only treatment and the lowest weight of 15\%, so that the structure of the framework and its 25/20/20/20/15 weighting are anchored in field evidence gathered from the sector itself rather than in intuition or in a transposition of foreign priorities. The weakest observed capability, governance, receives the heaviest weight; the dimension that can only be observed indirectly, skills and culture, receives the lightest. The weighting is thus a reasoned, evidence-informed choice rather than an arbitrary or stylistic one; as the construct-validity discussion below acknowledges, it was not tested against alternative weightings and should be read as defensible rather than provably optimal.

The fourth element is the depth of the scoring rubrics. The framework does not stop at naming 47 indicators; each indicator carries five explicit level descriptions, which amounts to 235 distinct evaluation criteria that were individually written and calibrated. This is the step that most maturity models omit and it is precisely what converts a maturity concept into an auditable instrument: an assessor confronted with an artifact does not interpret a vague label but matches the artifact against a written criterion.

The fifth element is the anti-gaming and regulatory rigor built into the methodology itself. The evidence cap, under which any indicator scored above two out of five without a documented artifact is capped at two, operationalizes the strategy versus formalization principle, according to which possessing a data strategy is not sufficient unless it is documented, approved, dated, and accessible. The proxy-only rule for the fifth dimension prevents unverifiable cultural claims from inflating the index, and the skip rule, bounded at twenty percent of the indicators per dimension and inapplicable to the thirteen non-skippable regulatory indicators, ensures that the regulatory core of the assessment can never be evaded. These rules were designed during the conceptual phase, before implementation, as properties of the method rather than as afterthoughts of the code.

The final element is method discipline. The conceptual phases were run as a structured consulting engagement: a four-step iterative method leading from benchmark through hybrid design and evidence-based scoring to tool development, a documented 24-week work breakdown covering the project from framing to closure, a stakeholder mapping formalized in an interest and influence matrix to plan engagement with each actor, and a formal SWOT analysis of the benchmark that explicitly weighed the option of adopting an existing framework against designing a bespoke one before the design decision was taken. Taken together, these six elements show that the pre-development work was not preparatory in the weak sense but constitutive: the defensibility of the DataPilot scores discussed throughout this chapter is inherited directly from the rigor of the phases that preceded the tool.

If a single thesis runs through the interpretation offered in this chapter, it is that the instrument is most defensible precisely where it is constrained. Each of the properties defended above is produced not by an added capability but by a deliberate restriction. The evidence cap bounds optimism: no claim can raise a score above two unless a documented artifact supports it. The skip ceiling bounds omission: at most two indicators may be set aside in Governance, two in Data Quality, and one in each of the remaining three dimensions, so a score cannot be reshaped by discarding inconvenient questions. The non-skippable status of the thirteen regulatory indicators bounds evasion: the regulatory core of the assessment can never be removed from the computation. The explicit proxy label and the fifteen percent weight bound the least observable dimension, so that the one part of the construct that cannot be verified directly is also the part with the least influence on the index. And the renormalization of the weights over the answered dimensions bounds the effect of an unanswered dimension, so that an incomplete assessment yields a proportionate figure rather than a distorted one. In each case the constraint is what converts a self assessment into a measurement: the score is trustworthy exactly where the method refuses to let it float, and the threats examined below are correspondingly concentrated where no such constraint can reach, namely in the honesty and quality of the evidence itself. The same logic carries into the fourth iteration, where the platform is defensible for the same reason the score is: the role hierarchy, the rule that every invitation creates a role exactly one rank below the inviter, bank-scoped tenant isolation, and the immutability of submissions are restrictions of exactly the same kind, and together they convert a shared platform into a defensible system of record. The confidentiality argument of the single-user milestone, which rested on keeping every response in the assessor's own browser, is thereby superseded for everything except in-progress solo answers: what now protects a bank is not where its data sits but the isolation the database itself enforces on every row, and that protection, too, is a constraint rather than a capability.

\subsection{Threats to Validity}

The validity of the study is examined along the four standard validity dimensions together with a researcher-bias threat, and the mitigations applied are described for each.

\subsubsection{Construct Validity}

Construct validity concerns whether the framework measures what it claims to measure, namely data maturity. The principal threat is that the choice of dimensions, sub-dimensions, and indicators could fail to capture the full construct. This threat was mitigated by deriving the structure from five established frameworks rather than from a single source or from intuition, by anchoring the regulatory indicators directly in the text of the BCT Circulaire and BCBS 239, and by attaching to each indicator an explicit five level rubric that defines precisely what is being measured. The multidimensional decomposition into five dimensions, twelve sub-dimensions, and 47 indicators is itself a construct validity mechanism: by breaking a broad and contested concept into narrow, separately observable questions, each with its own rubric, the framework reduces the risk that a single ambiguous item stands in for an entire capability. The residual threat concerns the fifth dimension, skills and culture, which cannot be observed directly and is assessed through proxy indicators such as training budget, headcount ratio, and certifications. While these proxies are objective and auditable, they measure investment in data culture rather than the culture itself, which is an acknowledged approximation. This limitation was deliberately bounded by assigning the dimension the smallest weight, 15\%, so that the approximation it introduces has the least influence on the global index, and by labeling the dimension explicitly as proxy based so that its interpretation is never mistaken for a direct measurement of culture. A related concession concerns the weighting itself: the 25\%, 20\%, 20\%, 20\%, and 15\% allocation is justified by the diagnostic evidence and by the regulatory centrality of governance, but it was not comparatively tested against alternative weightings, so it should be regarded as defensible rather than provably optimal, with the pilot assessments described in the future work as the natural setting for its calibration.

\subsubsection{Internal Validity}

Internal validity concerns whether the conclusions drawn from the assessment follow correctly from the data. The main threat is score inflation through declarative self assessment, the perception versus reality gap discussed above. This was mitigated by the evidence requirement and the evidence cap, which prevent any unverified claim from contributing a score above two. A second threat is the consistency of the aggregation: an error in the propagation from indicator to global score would invalidate the result. This was mitigated by validating the aggregation chain for consistency, confirming that a change at the indicator level propagates correctly through its sub-dimension and dimension to the GMI, and that the dimension weights sum to one hundred percent. A third threat, specific to the handling of missing responses, is that skipping or omitting indicators could distort a dimension score by silently changing its denominator. This was mitigated by two design rules working together: skipped and unanswered indicators are excluded from the averages rather than scored as zero, so an omission neither inflates nor deflates the result, and the number of skips is capped at twenty percent of the indicators in each dimension, which prevents a respondent from reshaping a score by discarding inconvenient questions. The renormalization of the global index over the weight of the answered dimensions applies the same principle at the top level, ensuring that a partially completed assessment still yields a proportionate and interpretable figure rather than one biased by the absence of a dimension. This renormalization is intended for a genuinely partial assessment, and it exposes one control that the delivered platform does not yet enforce: because an entirely unanswered dimension drops out of the weighted sum, a group coordinator who left the weakest dimension unmapped could raise the reported index by omission rather than by improvement. A finalized group submission should therefore be required to map all five dimensions, so that renormalization keeps a partial solo assessment meaningful without becoming a route to omit a dimension from an official submission; enforcing full dimension coverage at finalization is identified as a hardening step for future work. The non-skippable status of the thirteen regulatory indicators further protects internal validity, because the items most consequential for the conclusions can never be removed from the computation.

\subsubsection{External Validity}

External validity concerns the generalizability of the findings. The diagnostic is based on a survey of 44 banking professionals, which provides indicative rather than statistically exhaustive coverage of the sector, so the headline percentages should be read as directional evidence rather than precise population parameters. It is important to separate two distinct claims to generalizability here, because they carry different levels of risk. The first is the generalizability of the sector diagnostic, which is genuinely limited by the sample and is therefore presented as directional. The second is the generalizability of the framework as a measurement instrument, which does not depend on the survey at all: the framework is designed to be applied institution by institution, and its validity for a given bank rests on the evidence collected for that bank rather than on the survey sample. Confusing these two would overstate the risk, since a limited sample constrains what can be said about the sector but says nothing about whether the instrument measures a single institution correctly. The framework is specific to the banking sector and to the Tunisian regulatory environment by design, and its transfer to insurance or to other jurisdictions would require re-anchoring the regulatory indicators to the relevant texts. This bounded scope is a deliberate design decision rather than an oversight: the value of embedding thirteen indicators drawn from a specific national circular is inseparable from the fact that they are specific, and a version aimed at another sector or jurisdiction would substitute the corresponding regulatory mapping while retaining the same aggregation and evidence logic, which are domain independent.

\subsubsection{Conclusion Validity}

Conclusion validity concerns whether the relationship between the assessment and the maturity it reports is reliable and reproducible. Because the scoring methodology is fully deterministic, two assessors who enter the same raw scores and the same evidence will, by construction, arrive at the same GMI. This is computational reproducibility, and it is what allows a change in the index over time to be attributed to a real change in capability rather than to a change of assessor or of tool, which makes the figure usable as a tracked management indicator. It should not, however, be mistaken for inter-rater reliability in the stronger sense: two assessors may still read the same rubric and judge the same artifact differently when assigning the raw one-to-five score, so identical outputs are guaranteed only for identical inputs. Whether the instrument is reliable in that stronger sense, that is, whether independent assessors would produce the same inputs for the same bank, was not measured in this work and is identified as future work. The per-indicator rubric narrows this interpretive latitude at the point of scoring, so that the mapping from an observed artifact to a level is as unambiguous as the design can make it, but it is a mitigation rather than a proof of reliability. The principal residual threat is the quality and honesty of the evidence supplied, which the tool cannot verify autonomously and which remains the responsibility of the assessor. This threat is contained rather than eliminated: because every score above two must be justified by a named artifact that is retained in the exportable evidence package, a subsequent review or a regulatory inspection can re-examine the same artifacts and confirm or overturn the score, which places the honesty of the evidence under external check even though the tool itself cannot adjudicate it. The platform delivered in the fourth iteration strengthens this reliability further, because every answer in a group assessment records who supplied it and every finalized result is frozen as an immutable submission, so that neither the provenance nor the content of a reported figure can drift after the fact.

The word auditable deserves scrutiny, because it is the headline adjective of the methodological contribution and the exact basis of the claim should be stated. Auditability is here established by construction rather than by a field audit: the aggregation chain is deterministic, every score above two is tied to a named artifact, and the complete set of responses, evidence references, and results is exportable as a single package that an inspector could re-examine. What the evaluation demonstrates, having been conducted on simulated data, is therefore that an inspection could retrace every figure in the index back to its evidence, not that an inspection has done so. The claim is a property of the design, strongly evidenced at that level, and its confirmation under the conditions of a real supervisory review is deliberately deferred to the pilot assessments identified in the future work. Stating the boundary this precisely is what keeps the adjective honest: auditable here means built so that an audit would succeed, not already audited.

\subsubsection{Researcher Bias}

A threat that cuts across the four dimensions above is that a single author designed the framework, built the tool, seeded the demonstration data, and evaluated the results. This concentration of roles is a genuine researcher-bias threat: no independent party challenged the construct, scored a bank against the rubric, or judged whether an artifact was sufficient. Three factors contain, without eliminating, this threat. First, the instrument is deliberately auditable, so that an independent reviewer can retrace every displayed figure from the stored answers through the published rules and confirm or overturn it, which places the results under external check even though the evaluation itself was not independent. Second, the limitations of the evaluation, its reliance on seeded demonstration data and the absence of a field pilot or an automated test suite, are stated openly rather than hidden. Third, an independent evaluation, in which practitioners other than the author score real institutions and an inter-rater reliability study is carried out, is identified as the primary direction for future validation. Until such an evaluation is conducted, the results should be read as a demonstration of a faithful and internally consistent instrument rather than as an externally validated measurement.

\subsection{Conclusion}

This chapter interpreted the maturity deficit of the Tunisian banking sector as a coherent and self reinforcing pattern rooted in the governance dimension, articulated the theoretical and managerial value of the framework and the DataPilot tool, positioned the contribution relative to the existing frameworks, made explicit the rigor of the pre-development work from which the artifacts derive their defensibility, and examined the threats to validity together with their mitigations. The next chapter concludes the report and outlines the directions for future work.
