%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Abstract                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{\hfill Abstract \hfill}

\addcontentsline{toc}{section}{Abstract}

\vspace{1.4cm}
\large
\noindent
The Tunisian banking sector is subject to growing regulatory pressure on data management, driven by the Banque Centrale de Tunisie Circulaire N\textdegree{}2025-08 and by the international standard BCBS 239 (risk data aggregation standard), yet it lacks a standardized means of measuring and steering its data maturity, meaning how well it governs, controls, and exploits its data as a strategic asset. This project, conducted within the Data \& Analytics practice of EY Advisory Tunisia, addresses that gap by designing a data maturity assessment framework adapted to the banking sector and by developing an interactive tool that operationalizes it. The work follows a Design Science Research methodology for the design and validation of the framework, combined with Agile principles for the iterative development of the tool.

\noindent
A benchmark that screened fifteen international frameworks and retained five, namely DAMA-DMBOK, DCAM, CMMI-DMM, TDWI, and Gartner, together with a diagnostic survey of 44 banking professionals, informs the design. The resulting hybrid framework is structured around five dimensions, namely Governance, Data Quality, Architecture and Access, Analytics and Tools, and Skills and Culture, together with twelve sub-dimensions and 47 evidence based indicators, of which thirteen are non-skippable regulatory indicators tied to the BCT Circulaire N\textdegree{}2025-08 and the principles of BCBS 239. A weighted and renormalized scoring model, protected by an evidence cap (a rule that lowers an unsupported high score) that prevents over-optimistic self-assessment, produces an auditable Global Maturity Index on a one to five scale.

\noindent
The framework is operationalized through DataPilot, an interactive steering tool that delivers automated scoring, visual dashboards, gap analysis, a regulatory compliance view, and a phased improvement roadmap. DataPilot is delivered as a multi-tenant platform, meaning each bank is a separate tenant whose data no other bank can see. EY administers the EY master template and each bank receives its own copy, while four roles (EY Admin, Super Admin, Admin, Analyst) collaborate on solo and group assessments under database-enforced tenant isolation. Together, the framework and the tool provide Tunisian banks with a rigorous and actionable instrument to assess their data maturity and to support their digital transformation.

\vspace{0.6cm}
\noindent
\textbf{Keywords: Data maturity, Maturity assessment framework, Data governance, Banking sector, BCBS 239, Digital transformation, Multi-tenant platform, Web development, Software engineering, Design Science Research}
